---
title: "p8105_hw2_daq2107"
author: "Diego A. Quintana Licona (daq2107)"
date: "2026-10-06"
output: github_document
---

## Loading necessary libraries for this assignment:

```{r libraries}
library(tidyverse)
```

######### realtive path just means do not set a working directory at all, right?

## Problem 1

To start, we will load the data. We first need to recognize that the 'NYC Transit' data is coming from a 'csv' file. So we need to use the 'read_csv' function (part of tidyverse package) to import our file and then use the 'janitor' package to clean the names of columns which, after inspecting the file, contains complicated characters like uppercase letters:

```{r loading_csv}
nyc_transit =
  read_csv(file = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           )

# We use the 'janitor' package to clean the names:

nyc_transit = janitor::clean_names(nyc_transit)
```
Next, as per the assignment instructions, we need to select specific variables to keep. We can use the 'head' function to take a look at the variable names and then use the 'select' function to choose the variables we wish to keep:

```{r selecting_variables}

head(nyc_transit) # Exploring the data
str(nyc_transit) # Exploring the data more thoroughly

#Choosing the variables we need per the assignment instructions:

select(nyc_transit, line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada)

```






## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
