---
title: "p8105_hw2_daq2107"
author: "Diego A. Quintana Licona (daq2107)"
date: "2026-10-06"
output: github_document
---

## Loading necessary libraries for this assignment:

```{r libraries}

library(tidyverse) # Also loads 'dplyr' and 'readr' for data manipulation and export
library(readxl) # For Problem 2
```

## Problem 1: NYC Transit Data

To start, we will load and clean the data. We first need to recognize that the _NYC Transit_ data is coming from a _CSV_ file. So we need to use the _read_csv()_ function (part of tidyverse package) to import our file and then use the 'janitor' package to clean the names of columns which, after inspecting the file, contains complicated characters like uppercase letters. Notice also in the code below we are setting up a relative path to our data.

Additionally, as per the assignment instructions, we need to select specific variables to keep. I used the _head()_ function to take a look at the variable names and then used the _select()_ function to choose the variables to keep. We can also convert the _entry_ variable into a logical variable using the _case_match()_ function (part of _dplyr_ package in _tidyverse_) on the same code using the pipe operator:

```{r loading_and_cleaning}

nyc_transit_cleaned =
  read_csv(file = "./HW2_Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  select( # Choosing the variables we need per the assignment instructions:
    line, station_name, station_latitude, station_longitude, route1, route2, route3, route4,
    route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada) %>%
  mutate( # We use mutate since we are modifying the data set
    entry = case_match( # Converting YES/NO characters to logical type.
      entry,
      "YES" ~ TRUE, # Matching 'YES' to 'TRUE'
      "NO" ~ FALSE) # Matching 'NO' to 'FALSE'
  )

# Now, let's look at the newly modified data frame:

head(nyc_transit_cleaned, 4)

```

**Dataset Description:** The newly modified dataset is comprised of `r nrow(nyc_transit_cleaned)` observations and the `r ncol(nyc_transit_cleaned)` variables we selected above. The variables contain information on station line, the stations' name, latitude, longitude, as well as the route numbers. The dataset also contains some other infrastructure and managerial information, such as ADA compliance, staffing status, ticket vending, entrance type, among others.

We reduced the variable number from 32 to `r ncol(nyc_transit_cleaned)` variables using the _select()_ function, and then mutated the _entry_ variable to change it from a character variable to a logical variable. I would not consider the data to be tidy yet; we removed unnecessary variables for our analysis, coded for missing values, and have each row refer to a different observation and each column refer to a different measurement/variable, but we still need to manage our routes.

**How many distinct stations are there?**

To answer this question, we can use the _distinct()_ function (also part of the dplyr package).

```{r unique_stations}

distinct(nyc_transit_cleaned, station_name, line) %>% # Number of distinct stations
  nrow()
```

**Answer:** From the output, we can see that there are `r nrow(distinct(nyc_transit_cleaned, station_name, line))` distinct stations.

**How many stations are ADA compliant?**

For this, we consider that the ADA compliance variable is logical. Therefore, we can filter for the values of _ada_ that are TRUE, and then count the number of rows:

```{r ada_compliant}

nyc_transit_cleaned %>% 
  filter(ada == TRUE) %>% # Filtering for ADA compliant stations
  distinct(station_name, line) %>% # Running on unique stations
  nrow()

```

**Answer:** From the output, we can see that there are `r nrow(distinct(filter(nyc_transit_cleaned, ada == TRUE), line, station_name))` distinct ADA compliant stations.

**What proportion of station entrances / exits without vending allow entrance?**

We need to remember that we transformed the _entry_ variable into a logical variable (TRUE/FALSE), which, as we have seen in the previous Homework, can be counted in 1 and 0, respectively. Therefore, we can just take an average to get the proportion of entrances and exits after we filter for no vending.

```{r entrance_exit_prop}

nyc_transit_cleaned %>% 
  filter(vending == "NO") %>% 
  # I know the following syntax from a class I took last semester:
  summarise(prop_entrance = mean(entry, na.rm = TRUE))

```
**Answer:** From the output, we see that the proportion of proportion of entrances and exits after we filter for no vending is `r nyc_transit_cleaned %>% filter(vending == "NO") %>% pull(entry) %>% mean(na.rm = TRUE) %>% round(2)`.

**Reformatting Route Number & Route Name**

Now, we can reformat our data so that the route number and the route name are distinct variables. To do this, we can use the _pivot_longer()_ function. Following the example covered in lecture:

```{r route_name_and_number}

# I noticed that some route variables are character variables and others are numeric variables,
# so I am converting them all to characters first using mutate before using pivot_longer:

nyc_transit_tidier =
  nyc_transit_cleaned %>%  # We use the cleaned data frame
  mutate(route1 = as.character(route1),
         route2 = as.character(route2),
         route3 = as.character(route3),
         route4 = as.character(route4),
         route5 = as.character(route5),
         route6 = as.character(route6),
         route7 = as.character(route7),
         route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)) %>% 
  pivot_longer(
    route1:route11, # This takes all the route columns
    names_to = "route_number", # Making distinct route number
    names_prefix = "route", # Removing the common prefix
    values_to = "route_name") # Making distinct route name

# Let's visualize:

head(nyc_transit_tidier, 4) # First four elements for tidiness

```

**How many distinct stations serve the A train?**

Now that we have created distinct route names, we can simply use the _filter()_ function to get a count of the stations serving the A train:

```{r serves_A_train}

serve_A_train =
nyc_transit_tidier %>% 
  filter(route_name == "A") %>% # Filtering for A train
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

serve_A_train

```

**Answer:** From the output, we see that there are `r serve_A_train` distinct stations serving the A train.

**Of the stations that serve the A train, how many are ADA compliant?**

To answer this question we can run the same code as above but filtering for both A train _and_  ADA compliance:

```{r ADA_and_A_train}

ada_and_A =
nyc_transit_tidier %>% 
  filter(route_name == "A" & ada == TRUE) %>% # Filtering for A train and ADA compliance
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

ada_and_A

```

**Answer:** From the output, we see that there are `r ada_and_A` distinct stations that both serve the A train _and_ are ADA compliant.

## Problem 2: The Trash Wheels

The _Mr. Trash Wheel_ dataset is an Excel file, and therefore, we need to use the _readxl_ package to import our data. We can use the _janitor_ package to make sure we have reasonable variable names (like we did in Problem 1), use _drop_na()_ to omit missing observations for our _dumpster_ variable and then use _round_ and _as.integer_ within _mutate()_ to make sure we round the number of sports balls to the nearest integer.

```{r importing_cleaning_mrtrashwheel}

# We can use the 'sheet' argument to specify the sheet we want and the 'range' argument to help
# us omit non-data entries.

trash_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Mr. Trash Wheel",
                            range = "A2:N742") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) %>% # Omitting missing values
  mutate(
    sports_balls = as.integer(round(sports_balls)) # Converting to integer
  )

#Inspecting the cleaned data frame:

head(trash_wheel_df, 4) 

```

We can do a similar tidying process for the _Professor Trash Wheel_ and _Gwynnda the Good Wheel of the West_ sheets in the dataset:

```{r professor_and_gwynda}

# For 'Professor Trash Wheel':

prof_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Professor Trash Wheel",
                            range = "A2:M139") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) # Dropping missing values

head(prof_wheel_df, 4) #Inspecting the cleaned data frame

# For 'Gwynnda the Good Wheel of the W':

gwynnda_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Gwynnda the Good Wheel of the W",
                            range = "A2:L394") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) # Dropping missing values

head(gwynnda_wheel_df, 4) #Inspecting the cleaned data frame


```

Now that we have created three separate data frames for the specified Excel sheets, we can combine them. To do this, we can use the _bind_rows()_ function covered during the _Tidy Data_ lecture. Note that, before we combine, we first need to add a variable in each data frame so we know where the data is coming from. We can do this using _mutate()_.

```{r combining_trash_sheets}

# On my first attempt, I got an error that in one data frame 'year' and 'plastic_bottles' were
# characters, and in another they were doubles (numbers). So I added an as.numeric conversion
# within mutate to resolve the issue:

trash_wheel_df =
  trash_wheel_df %>% 
  mutate(source_sheet = "Mr. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

prof_wheel_df =
  prof_wheel_df %>% 
  mutate(source_sheet = "Prof. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

gwynnda_wheel_df =
  gwynnda_wheel_df %>% 
  mutate(source_sheet = "Gwynnda the Good Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

# Now we can use 'bind_rows' to combine:

combined_trash_df =
  bind_rows(trash_wheel_df, prof_wheel_df, gwynnda_wheel_df)

# Inspecting the newly combined data frame:

head(combined_trash_df, 4)
```

**Data Description:** We now have a combined data frame! It is comprised of `r ncol(combined_trash_df)` variables and `r nrow(combined_trash_df)` observations. Some key variables included in the data frame are the tons of trash collected by each dumpster, the amount of homes powered, and a breakdown of the number of plastic items collected, such as bottles, bags and cigarette butts.

Interestingly, across the three trash collectors, on average, they have collected approximately `r round(mean(pull(combined_trash_df, weight_tons), na.rm = TRUE), 2)` tons of waste, `r format(round(mean(pull(combined_trash_df, cigarette_butts), na.rm = TRUE), 2), scientific = FALSE)` cigarette butts, and `r round(mean(pull(combined_trash_df, plastic_bags), na.rm = TRUE), 2)` plastic bags between the years `r min(pull(combined_trash_df, year), na.rm = TRUE)` and `r max(pull(combined_trash_df, year), na.rm = TRUE)`.

**For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?**

To answer these questions we can make use of the _filter()_ function to delimit by the specified parameters:

```{r specific_trash_questions}

# For Professor Trash Wheel:

total_weight_prof = 
  combined_trash_df %>% 
  filter(
    source_sheet == "Prof. Trash Wheel"
  ) %>% 
  pull(weight_tons) %>% 
  sum(na.rm = TRUE)

total_weight_prof

# For Gwynnda the Good Wheel:

total_cigs_Gwynnda =
  combined_trash_df %>% 
  filter(
    source_sheet == "Gwynnda the Good Wheel" & 
            year == 2022 &
           month == "June"
  ) %>% 
  pull(cigarette_butts) %>% 
  sum(na.rm = TRUE)

total_cigs_Gwynnda

```

**Answer:** From the output, we can see that _Professor Trash Wheel_ collected a total of `r total_weight_prof` tons of trash. Additionally, _Gwynnda the Good Wheel of the West_ collected a total of `r format(round(total_cigs_Gwynnda, 2), scientific = FALSE)` cigarette butts in June 2022 alone.

## Problem 3: Alzheimer’s Disease (AD) & Mild Cognitive Impairment (MCI)

**Baseline**

Per usual, let's first import and clean the data.

```{r MCI_baseline_import}

# After inspecting the CSV, we realize that the first row is just general descriptors, so can 
# use the 'skip' argument when importing the file to skip this first row.

# Amount of people recruited:

baseline_df =
  read_csv(file = "./HW2_Data/data_mci/MCI_baseline.csv",
           skip = 1, # Skipping first row
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  mutate(
    sex = case_match(
      sex,
      1 ~ "Male", # Matching '1' to 'Male'
      0 ~ "Female"), # Matching '0' to 'Female'
    apoe4 = case_match(
      apoe4,
      1 ~ "Carrier", # Matching '1' to 'Carrier'
      0 ~ "Non-Carrier") # Matching '0' to 'Non-Carrier'
    ) %>% 
  filter(
    age_at_onset > current_age | is.na(age_at_onset)
    )
 
# Inspecting the data frame:

head(baseline_df, 4) 

```
**Import Process and Data Frame Description:**  Since the data is coming from a CSV file, we used _read_csv()_ in the same fashion as in Problem 1. We will also need to use the _skip_ argument within _read_csv_ given that the first row is plain descriptions of the variables. As we have previously done, we use the _janitor_ package to make sure our variable names are readable. We then use _case_match_ like we did in previous problems to make sure that the _sex_ and _apoe4_ variables are recoded as categorical rather than numerical variables.

Finally, we need to make sure that the observations included comply with the inclusion criteria where they must not have had MCI by the time of baseline. To do this, we filter so we keep observations where the _age_at_onset_ is greater than the _current_age_, retaining observations with NAs.

The resulting data frame for the baseline demographic metrics is comprised of `r nrow(baseline_df)` observations and `r ncol(baseline_df)` variables, where each unit of observation corresponds to one study participant at baseline. The data frame contains information such as baseline age, sex, APOE4 carrier status, and age at MCI onset.

**How many participants were recruited? Of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?**

```{r baseline_analysis}

# Number of recruits:

pts_recruited = 
  nrow(baseline_df) 

pts_recruited

# Develop MCI:

develop_mci =
  baseline_df %>% 
  drop_na(age_at_onset) %>% 
  nrow()

develop_mci

# Baseline age:

baseline_age =
  baseline_df %>% 
  pull(current_age) %>% 
  mean(na.rm = TRUE)

baseline_age

# Woman carrying APOE4:

women_carrier =
  baseline_df %>% 
  filter(
    sex == "Female"
  ) %>% 
  summarize(
    proportion_women = 
      mean(apoe4 == "Carrier", na.rm = TRUE))

women_carrier

```
**Answer:** From the outputs, we can see that there were a total of `r pts_recruited` participants recruited, of which `r develop_mci` developed MCI during the follow-up period. Moreover, the average baseline age for the study participants is `r round(baseline_age, 2)` and, from the participants who identify as female, `r women_carrier * 100`% of them are carriers of the APOE4 gene variant (proportion = `r women_carrier`).

**Longitudinally observed biomarker values**

We can clean the longitudinal amyloid biomarker data in a similar fashion as for the baseline data.

```{r MCI_amyloid_import}

# Upon inspection, we notice that the amyloid measurements (time_) are saved as characters. So we modify to
# numeric since they are years.

amyloid_df =
  read_csv(file = "./HW2_Data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  mutate( # Transforming to numeric types
    baseline = as.numeric(baseline),
    time_2 = as.numeric(time_2),
    time_4 = as.numeric(time_4),
    time_6 = as.numeric(time_6),
    time_8 = as.numeric(time_8),
    ) %>% 
   rename(
     id = study_id) %>% # We do this for the join later (see below)
  pivot_longer( # Transforming into long format
    baseline:time_8, # This takes all the time columns
    names_to = "years_since_baseline", # Making distinct time variable
    names_prefix = "time_", # Removing the common prefix
    values_to = "amyloid") %>%
  mutate(
    years_since_baseline = 
      replace(years_since_baseline,
              years_since_baseline == "baseline", 0),
    years_since_baseline = as.numeric(years_since_baseline)) #Making it numeric

# Inspecting the data:

head(amyloid_df, 4)

```

**Import Process and Data Frame Description:** Similar to the baseline data, the longitudinal data is coming from a CSV file, so we use _read_csv()_ to import. We will again need to use the _skip_ argument within _read_csv_ given that the first row is descriptions of the variables. As we have previously done, we use the _janitor_ package to make sure our variable names are readable. We then use _mutate()_ and _as.numeric()_ to convert the variables representing the amyloid measurement time since baseline into numeric variables. We then proceed to rename the _study_id_ variable into just _id_ so it matches our baseline data frame, and use _pivot_longer()_ to transform our data into a long format while creating the new variable _years_since_baseline_ to identify the timing of each measurement. We also created the _amyloid_ variable to contain the corresponding amyloid 42/40 ratio. We finally had to code our baseline variable as zero years to ensure that the years count remains consistent.

The resulting data frame for the longitudinal amyloid data is comprised of `r nrow(amyloid_df)` observations and `r ncol(amyloid_df)` variables, where each unit of observation corresponds to one amyloid measurement for each participant during the follow-up period. Therefore, multiple rows may correspond to the same participant. The data frame contains variables for participant ID, time since baseline, and amyloid ratio.

**Joining datasets**

Since now both data frames have a common _id_ variable, we can use an _anti_join()_ to identify participants missing in one data set but present in the other, and then _inner_join_ to merge them.

```{r joining_amyloid_mci}

# Finds people in baseline but not amyloid:

anti_join(baseline_df, amyloid_df) %>% 
  distinct(id)

# Finds people in amyloid but not in baseline:

anti_join(amyloid_df, baseline_df) %>% 
  distinct(id)
  
# Merging!

merged_mci_df =
  inner_join(baseline_df, amyloid_df)

# Inspecting our merge:

head(merged_mci_df, 4)

```

**Answer:** From our anti-join operations, we can see that there are `r anti_join(baseline_df, amyloid_df, by = "id") %>% distinct(id) %>% nrow()` participants who are in the baseline data frame but not the longitudinal amyloid data frame, while there are `r anti_join(amyloid_df, baseline_df, by = "id") %>% distinct(id) %>% nrow() ` participants who are in the longitudinal amyloid data frame and not in the baseline data frame.

The combined data set is now comprised of `r nrow(merged_mci_df)` observations and `r ncol(merged_mci_df)` variables. Each observation unit corresponds to participants who were in both data frames (both baseline and follow-up) at a particular follow-up. The variables now include both demographics (current age, sex, APOE4 carrier status, age at MCI onset) and amyloid tracking data in years since baseline.

**We can finally export our merged data frame to our data directory**

```{r exporting}

write_csv(merged_mci_df, 
          "./HW2_Data/merged_mci_df.csv")

```

## The End :) Have a wonderful day!
