---
title: "p8105_hw2_daq2107"
author: "Diego A. Quintana Licona (daq2107)"
date: "2026-10-06"
output: github_document
---

##### do I need to load dplyr? or included in tidyverse??


## Loading necessary libraries for this assignment:

```{r libraries}

library(tidyverse)
library(dplyr)
library(readxl) # For Problem 2
```

######### realtive path just means do not set a working directory at all? Or making folder only?

## Problem 1

To start, we will load the data. We first need to recognize that the 'NYC Transit' data is coming from a 'csv' file. So we need to use the 'read_csv' function (part of tidyverse package) to import our file and then use the 'janitor' package to clean the names of columns which, after inspecting the file, contains complicated characters like uppercase letters:

```{r loading_csv}

# We use the 'janitor' package after reading the file to clean the names:
nyc_transit =
  read_csv(file = "./HW2_Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names()
```
Next, as per the assignment instructions, we need to select specific variables to keep. We can use the 'head' function to take a look at the variable names and then use the 'select' function to choose the variables we wish to keep. We can also convert the 'entry' variable into a logical variable using the 'case_match' function (part of dplyr package) on the same code using the pipe operator:

```{r selecting_variables}

head(nyc_transit) # Exploring the data

# Choosing the variables we need per the assignment instructions:

nyc_transit_cleaned = nyc_transit %>%
  select(
    line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada) %>% 
  mutate( # We use mutate since we are modifying the data set
    entry = case_match( # Converting YES/NO characters to logical type.
      entry,
      "YES" ~ TRUE, # Matching 'YES' to 'TRUE'
      "NO" ~ FALSE) # Matching 'NO' to 'FALSE'
  )

# Now, lets look at the newly modified data frame:

str(nyc_transit_cleaned)
nrow(nyc_transit_cleaned)
ncol(nyc_transit_cleaned)

```
#### Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

##### how can we know if the data is tidy???????
############# maybe use inline R


**Dataset Description:** The newly modified dataset is comprised by 1868 observations and the 20 variables we selected above. The variables contain information on the transit division, the stations' name, latitude, and longitude, as well as the line names and route numbers. The dataset also contains some other infrastructure and managerial information, such as entrance type and location, what streets do exits lead to, ADA compliance, staffing status, ticket vending, among others.

We reduced the variable number from 32 to 20 variables using the 'select' function, and then mutated the 'entry' variable to change it from a character variable to a logic variable. I would now consider the data to be tidy, as we removed unnecessary variables for our analysis and coded for missing values.

**How many distinct stations are there?**

To answer this question, we can use the 'distinct' function (also part of the dplyr package)

```{r unique_stations}

distinct(nyc_transit_cleaned, station_name, line) %>% # Number of distinct stations
  nrow()
```

From the output, we can see that there are 465 distinct stations.

**How many stations are ADA compliant?**

For this, we consider that the ADA compliance variable is logical. Therefore, we can filter for the values of _ada_ that are TRUE, and then count the number of rows:

```{r ada_compliant}

nyc_transit_cleaned %>% 
  filter(ada == TRUE) %>% # Filtering for ADA compliant stations
  distinct(station_name, line) %>% # Running on unique stations
  nrow()

```

From the output, we can see that there are 84 distinct ADA compliant stations.

**What proportion of station entrances / exits without vending allow entrance?**

We need to remember that we transformed the 'entry' variable into a logical variable (TRUE/FALSE),
which, as we have seen in the previous Homework, can be counted in 1 and 0, respectively. Therefore, we can just take an average to get the proportion of entrances and exits after we filter for no vending:

```{r entrance_exit_prop}

nyc_transit_cleaned %>% 
  filter(vending == "NO") %>% 
  # I know the following syntax from a class I took last semester:
  summarise(prop_entrance = mean(entry, na.rm = TRUE))

```
From the output, we see that the proportion of entrance to exits in unique stations is 37.7%

*Reformatting Route Number & Route Name**

Now, we can reformat our data so that the route number and the route name are distinct variables. To do this, we can use the _pivot_longer_ function. Following the example covered in lecture:

```{r route_name_and_number}

# I noticed that some route variables are character variables and others are numeric variables, so I am converting them all to characters first using mutate before using pivot_longer:

nyc_transit_tidier =
  nyc_transit_cleaned %>%  # We use the cleaned data frame
  mutate(route1 = as.character(route1),
         route2 = as.character(route2),
         route3 = as.character(route3),
         route4 = as.character(route4),
         route5 = as.character(route5),
         route6 = as.character(route6),
         route7 = as.character(route7),
         route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)) %>% 
  pivot_longer(
    route1:route11, # This takes all the route columns
    names_to = "route_number", # Making distinct route number
    names_prefix = "route", # Removing the common prefix
    values_to = "route_name") # Making distinct route name

# Let's visualize:

nyc_transit_tidier

```

**How many distinct stations serve the A train? **

Now that we have created distinct route names, we can simply use the _filter_ function to get a count of the stations serving the A train:

```{r serves_A_train}

nyc_transit_tidier %>% 
  filter(route_name == "A") %>% # Filtering for A train
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

```

############# NYC transit says it is 47, need to check

From the output, we see that there are 60 distinct stations serving the A train.

**Of the stations that serve the A train, how many are ADA compliant?**

To answer this question we can run the same code as above but filtering for both A train _and_  ADA compliance:

```{r ADA_and_A_train}

nyc_transit_tidier %>% 
  filter(route_name == "A" & ada == TRUE) %>% # Filtering for A train and ADA compliance
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

```

From the output, we see that there are 17 distinct stations that both serve the A train _and_ are ADA compliant.

############## ask about .DS_Store file????


## Problem 2

The _Mr. Trash Wheel_ dataset is an Excel file, and therefore, we need to use the _readxl_ package to import our data:

############### should I put everything on the same pipe?
######### I know 'round' from another class, do I need to state that?

```{r importing_excel}

# We can use the 'sheet' argument to specify the sheet we want and the 'range' argument to help us omit non-data entries.

# We use the 'janitor' package to make sure we have reasonable variable names.

# We can also use the 'drop_na' function to omit missing observations for our 'dumpster' variable and then use 'round' and 'as.integer' within 'mutate' to make sure we round the number of sports ball to the nearest integer.

trash_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Mr. Trash Wheel",
                            range = "A2:N742") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )

trash_wheel_df #Inspecting the cleaned data frame

```

We can do a similar tidying process for the _Professor Trash Wheel_ and _Gwynda the Good Wheel of the West_ sheets in the dataset:

###### no sports balls for these right?

```{r professor_and_gwynda}

# For 'Professor Trash Wheel':

prof_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Professor Trash Wheel",
                            range = "A2:M139") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster)

prof_wheel_df #Inspecting the cleaned data frame

# For 'Gwynnda the Good Wheel of the W':

gwynnda_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Gwynnda the Good Wheel of the W",
                            range = "A2:L394") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster)

gwynnda_wheel_df #Inspecting the cleaned data frame


```

Now that we have created three separate data frames for the specified Excel sheets, we can combine them. To do this, we can use the _bind_rows_ function covered during the Tidy Data lecture:

```{r combining_trash_sheets}

# Before we combine, we first need to add a variable in each data frame so we know where the data is coming from. Additionally, on my first attempt, I got an error that in one data frame 'year' and 'plastic_bottles' were characters, and in another they were doubles (numbers). So I added an as.numeric conversion within mutate to resolve the issue:

trash_wheel_df =
  trash_wheel_df %>% 
  mutate(source_sheet = "Mr. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

prof_wheel_df =
  prof_wheel_df %>% 
  mutate(source_sheet = "Prof. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

gwynnda_wheel_df =
  gwynnda_wheel_df %>% 
  mutate(source_sheet = "Gwynnda the Good Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

# Now we can use 'bind_rows' to combine:

combined_trash_df =
  bind_rows(trash_wheel_df, prof_wheel_df, gwynnda_wheel_df)

# Inspecting the newly coombined data frame:

combined_trash_df
names(combined_trash_df)

```

**Data description:**

We now have a combined data frame! It is comprised of `r ncol(combined_trash_df)` variables and `r nrow(combined_trash_df)` observations. Some key variables included in the data frame are the tons of trash collected by each dumpster, the amount of homes powered, and a break down of the number of plastic items collected, such as bottles, bags and cigarette butts. Interestingly, across the three trash collectors, oon average, they have collected approximately `r mean(pull(combined_trash_df, weight_tons), na.rm = TRUE)` tons of waste, `r mean(pull(combined_trash_df, cigarette_butts), na.rm = TRUE)` cigarette butts, and `r mean(pull(combined_trash_df, plastic_bags), na.rm = TRUE)` plastic bags between the years `r min(pull(combined_trash_df, year), na.rm = TRUE)` and `r max(pull(combined_trash_df, year), na.rm = TRUE)`.

**For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?**

To answer these questions we can make use of the _filter_ function to delimit by the specified parameters:

```{r specific_trash_questions}

# For Professor Trash Wheel:

  combined_trash_df %>% 
  filter(
    source_sheet == "Prof. Trash Wheel"
  ) %>% 
  pull(weight_tons) %>% 
  sum(na.rm = TRUE)

# For Gwynnda the Good Wheel:

combined_trash_df %>% 
  filter(
    source_sheet == "Gwynnda the Good Wheel" & 
            year == 2022 &
           month == "June"
  ) %>% 
  pull(cigarette_butts) %>% 
  sum(na.rm = TRUE)


```

##### I dont thinnk I can use inline R code here??

**Answer:** From the output, we can see that _Professor Trash Wheel_ collected a total of 294.96 tons of trash. Additionally, _Gwynnda the Good Wheel_ collected a total of 18,120 cigarette butts in June, 2022 alone.

## Problem 3




