---
title: "p8105_hw2_daq2107"
author: "Diego A. Quintana Licona (daq2107)"
date: "2026-10-06"
output: github_document
---

## Loading necessary libraries for this assignment:

```{r libraries}

library(tidyverse) # Also loads 'dplyr'
library(readxl) # For Problem 2
```

## Problem 1: NYC Transit Data

To start, we will load and clean the data. We first need to recognize that the 'NYC Transit' data is coming from a 'csv' file. So we need to use the 'read_csv' function (part of tidyverse package) to import our file and then use the 'janitor' package to clean the names of columns which, after inspecting the file, contains complicated characters like uppercase letters. Notice also in the code below we are setting up a relative path to our data.

Additionally, as per the assignment instructions, we need to select specific variables to keep. I used the 'head' function to take a look at the variable names and then used the 'select' function to choose the variables to keep. We can also convert the 'entry' variable into a logical variable using the 'case_match' function (part of dplyr package) on the same code using the pipe operator:

```{r loading_and_cleaning}

nyc_transit_cleaned =
  read_csv(file = "./HW2_Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  select( # Choosing the variables we need per the assignment instructions:
    line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada) %>%
  mutate( # We use mutate since we are modifying the data set
    entry = case_match( # Converting YES/NO characters to logical type.
      entry,
      "YES" ~ TRUE, # Matching 'YES' to 'TRUE'
      "NO" ~ FALSE) # Matching 'NO' to 'FALSE'
  )

# Now, lets look at the newly modified data frame:

head(nyc_transit_cleaned, 4)

```

**Dataset Description:** The newly modified dataset is comprised by `r nrow(nyc_transit_df)` observations and the `r ncol(nyc_transit_df)` variables we selected above. The variables contain information on the transit division, the stations' name, latitude, and longitude, as well as the line names and route numbers. The dataset also contains some other infrastructure and managerial information, such as entrance type and location, what streets do exits lead to, ADA compliance, staffing status, ticket vending, among others.

We reduced the variable number from 32 to `r ncol(nyc_transit_df)` variables using the 'select' function, and then mutated the 'entry' variable to change it from a character variable to a logic variable. I would now consider the data to be tidy, as we removed unnecessary variables for our analysis, coded for missing values, and have each row refer to a different observation and each column refer to a different measurement/variable.

**How many distinct stations are there?**

To answer this question, we can use the _distinct()_ function (also part of the dplyr package).

```{r unique_stations}

distinct(nyc_transit_cleaned, station_name, line) %>% # Number of distinct stations
  nrow()
```

From the output, we can see that there are `r nrow(distinct(nyc_transit_cleaned, station_name, line))` distinct stations.

**How many stations are ADA compliant?**

For this, we consider that the ADA compliance variable is logical. Therefore, we can filter for the values of _ada_ that are TRUE, and then count the number of rows:

```{r ada_compliant}

nyc_transit_cleaned %>% 
  filter(ada == TRUE) %>% # Filtering for ADA compliant stations
  distinct(station_name, line) %>% # Running on unique stations
  nrow()

```

From the output, we can see that there are `r nrow(distinct(filter(nyc_transit_cleaned, ada == TRUE), line, station_name))` distinct ADA compliant stations.

**What proportion of station entrances / exits without vending allow entrance?**

We need to remember that we transformed the _entry_ variable into a logical variable (TRUE/FALSE), which, as we have seen in the previous Homework, can be counted in 1 and 0, respectively. Therefore, we can just take an average to get the proportion of entrances and exits after we filter for no vending.

```{r entrance_exit_prop}

nyc_transit_cleaned %>% 
  filter(vending == "NO") %>% 
  # I know the following syntax from a class I took last semester:
  summarise(prop_entrance = mean(entry, na.rm = TRUE))

```
From the output, we see that the proportion of entrance to exits in unique stations is `r nyc_transit_cleaned %>% filter(vending == "NO") %>% pull(entry) %>% mean(na.rm = TRUE)`.

*Reformatting Route Number & Route Name**

Now, we can reformat our data so that the route number and the route name are distinct variables. To do this, we can use the _pivot_longer()_ function. Following the example covered in lecture:

```{r route_name_and_number}

# I noticed that some route variables are character variables and others are numeric variables, so I am converting them all to characters first using mutate before using pivot_longer:

nyc_transit_tidier =
  nyc_transit_cleaned %>%  # We use the cleaned data frame
  mutate(route1 = as.character(route1),
         route2 = as.character(route2),
         route3 = as.character(route3),
         route4 = as.character(route4),
         route5 = as.character(route5),
         route6 = as.character(route6),
         route7 = as.character(route7),
         route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)) %>% 
  pivot_longer(
    route1:route11, # This takes all the route columns
    names_to = "route_number", # Making distinct route number
    names_prefix = "route", # Removing the common prefix
    values_to = "route_name") # Making distinct route name

# Let's visualize:

head(nyc_transit_tidier, 4) # First four elements for tidiness

```

**How many distinct stations serve the A train? **

Now that we have created distinct route names, we can simply use the _filter()_ function to get a count of the stations serving the A train:

```{r serves_A_train}

serve_A_train =
nyc_transit_tidier %>% 
  filter(route_name == "A") %>% # Filtering for A train
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

serve_A_train

```

############# NYC transit says it is 47, need to check

From the output, we see that there are `r serve_A_train` distinct stations serving the A train.

**Of the stations that serve the A train, how many are ADA compliant?**

To answer this question we can run the same code as above but filtering for both A train _and_  ADA compliance:

```{r ADA_and_A_train}

ada_and_A =
nyc_transit_tidier %>% 
  filter(route_name == "A" & ada == TRUE) %>% # Filtering for A train and ADA compliance
  distinct(station_name, line) %>% # Running on unique stations, like in previous question
  nrow() # Obtaining the number of observations

ada_and_A

```

From the output, we see that there are `r serve_A_train` distinct stations that both serve the A train _and_ are ADA compliant.

## Problem 2: The Trash Wheels

The _Mr. Trash Wheel_ dataset is an Excel file, and therefore, we need to use the _readxl_ package to import our data. We can use the _janitor_ package to make sure we have reasonable variable names (like we did in Problem 1), use _drop_na()_ to omit missing observations for our _dumpster_ variable and then use _round_ and _as.integer_ within _mutate()_ to make sure we round the number of sports ball to the nearest integer.

```{r importing_cleaning_mrtrashwheel}

# We can use the 'sheet' argument to specify the sheet we want and the 'range' argument to help us omit non-data entries.

trash_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Mr. Trash Wheel",
                            range = "A2:N742") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) %>% # Omitting missing values
  mutate(
    sports_balls = as.integer(round(sports_balls)) # Converting to integer
  )

#Inspecting the cleaned data frame:

head(trash_wheel_df, 4) 

```

We can do a similar tidying process for the _Professor Trash Wheel_ and _Gwynda the Good Wheel of the West_ sheets in the dataset:

```{r professor_and_gwynda}

# For 'Professor Trash Wheel':

prof_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Professor Trash Wheel",
                            range = "A2:M139") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) # Dropping missing values

prof_wheel_df #Inspecting the cleaned data frame

# For 'Gwynnda the Good Wheel of the W':

gwynnda_wheel_df = read_excel("./HW2_Data/202610 Trash Wheel Collection Data.xlsx",
                            sheet = "Gwynnda the Good Wheel of the W",
                            range = "A2:L394") %>% 
  janitor::clean_names() %>% # Cleaning names
  drop_na(dumpster) # Dropping missing values

gwynnda_wheel_df #Inspecting the cleaned data frame


```

Now that we have created three separate data frames for the specified Excel sheets, we can combine them. To do this, we can use the _bind_rows()_ function covered during the _Tidy Data_ lecture. Note that, before we combine, we first need to add a variable in each data frame so we know where the data is coming from. We can do this using _mutate()_.

```{r combining_trash_sheets}

# On my first attempt, I got an error that in one data frame 'year' and 'plastic_bottles' were characters, and in another they were doubles (numbers). So I added an as.numeric conversion within mutate to resolve the issue:

trash_wheel_df =
  trash_wheel_df %>% 
  mutate(source_sheet = "Mr. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

prof_wheel_df =
  prof_wheel_df %>% 
  mutate(source_sheet = "Prof. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

gwynnda_wheel_df =
  gwynnda_wheel_df %>% 
  mutate(source_sheet = "Gwynnda the Good Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))

# Now we can use 'bind_rows' to combine:

combined_trash_df =
  bind_rows(trash_wheel_df, prof_wheel_df, gwynnda_wheel_df)

# Inspecting the newly combined data frame:

combined_trash_df
names(combined_trash_df)

```

**Data description:**

We now have a combined data frame! It is comprised of `r ncol(combined_trash_df)` variables and `r nrow(combined_trash_df)` observations. Some key variables included in the data frame are the tons of trash collected by each dumpster, the amount of homes powered, and a break down of the number of plastic items collected, such as bottles, bags and cigarette butts.

Interestingly, across the three trash collectors, on average, they have collected approximately `r mean(pull(combined_trash_df, weight_tons), na.rm = TRUE)` tons of waste, `r mean(pull(combined_trash_df, cigarette_butts), na.rm = TRUE)` cigarette butts, and `r mean(pull(combined_trash_df, plastic_bags), na.rm = TRUE)` plastic bags between the years `r min(pull(combined_trash_df, year), na.rm = TRUE)` and `r max(pull(combined_trash_df, year), na.rm = TRUE)`.

**For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?**

To answer these questions we can make use of the _filter_ function to delimit by the specified parameters:

```{r specific_trash_questions}

# For Professor Trash Wheel:

  combined_trash_df %>% 
  filter(
    source_sheet == "Prof. Trash Wheel"
  ) %>% 
  pull(weight_tons) %>% 
  sum(na.rm = TRUE)

# For Gwynnda the Good Wheel:

combined_trash_df %>% 
  filter(
    source_sheet == "Gwynnda the Good Wheel" & 
            year == 2022 &
           month == "June"
  ) %>% 
  pull(cigarette_butts) %>% 
  sum(na.rm = TRUE)


```

##### I dont thinnk I can use inline R code here?? ASSIGN TO VARIABLE AND CALL N INLINE R

**Answer:** From the output, we can see that _Professor Trash Wheel_ collected a total of 294.96 tons of trash. Additionally, _Gwynnda the Good Wheel_ collected a total of 18,120 cigarette butts in June, 2022 alone.

## Problem 3: Alzheimer’s Disease (AD) & Mild Cognitive Impairment (MCI)

Let's first import the data. We see that these are CSV files, so we can use the same import process as in Problem 1:

```{r MCI_baseline_import}

# After inspecting the CSV, we realize that the first row is just general descriptors, so can use the 'skip' argument when importing the file to skip this first row.

# We can also make use of 'case_match' like we did in previous problems to make sure the sex and APOE4 variables are stored not as numerical variables:

# Amount of people recruited:

baseline_df =
  read_csv(file = "./HW2_Data/data_mci/MCI_baseline.csv",
           skip = 1,
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  mutate(
    sex = case_match(
      sex,
      1 ~ "Male", # Matching '1' to 'Male'
      0 ~ "Female"), # Matching '0' to 'Female'
    apoe4 = case_match(
      apoe4,
      1 ~ "Carrier", # Matching '1' to 'Carrier'
      0 ~ "Non-Carrier") # Matching '0' to 'Non-Carrier'
    ) %>% 
  filter(
    age_at_onset > current_age | is.na(age_at_onset)
    )
 


 #Inspecting the data frame

##### DROP THOSE THAT HAD MCI AT BASELINE. USE FILTER
##### HOW MANY DEVELOP MCI

str(baseline_df) #Inspecting the data frame


```
**Import Process and Data Frame Description:**  
Mention main steps of cleaning and import

**How many participants were recruited? Of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?**

```{r baseline_analysis}

nrow(baseline_df) # Number of recruits

### 479 recruits

# How many develop MCI:

baseline_w_mci =
  baseline_df %>% 
  drop_na(age_at_onset)

#### 93 develop MCI


women_carrier =
  baseline_df %>% 
  filter(
    sex == "Female" & apoe4 == "Carrier"
  )

### 63 women

```






######## HOW ELSE TO TIDY?

```{r MCI_amyloid_import}

# Upon inspection, we notice that the timepoints are saved as characters. So we modify to numeric since they are years.

amyloid_df =
  read_csv(file = "./HW2_Data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("","NA", ".") # We use this to make sure R understands these patterns as NA
           ) %>% 
janitor::clean_names() %>% # Cleaning names
  mutate(
    baseline = as.numeric(baseline),
    time_2 = as.numeric(time_2),
    time_4 = as.numeric(time_4),
    time_6 = as.numeric(time_6),
    time_8 = as.numeric(time_8),
    ) %>% 
   rename(
     id = study_id) %>% # We do this for the join later (see below)
  pivot_longer(
    baseline:time_8, # This takes all the time columns
    names_to = "years_since_baseline", # Making distinct time variable
    names_prefix = "time_", # Removing the common prefix
    values_to = "amyloid") %>%
  mutate(
    years_since_baseline = 
      replace(years_since_baseline,
              years_since_baseline == "baseline", 0))

head(baseline_df)

head(amyloid_df)

```


```{r joining_amyloid_mci}

# Since now both data frames have a common 'id' variable, we can use an antijoin to merge them. We need to run this process twice 

anti_join(baseline_df, amyloid_df) # Finds people in baseline but not amyloid (8 people)

anti_join(amyloid_df, baseline_df) # Finds people in amyloid but not in baseline (80 people)

merged_mci_df =
  inner_join(baseline_df, amyloid_df)

```



###### MERGE SSO EVERY ROW ISS AN APPOINTMENT WHEN SOMEONNW GOT MEASUREMENTS TAKEN
###### KEY TO MERGE -- SO WE USE A JOIN STUDY ID. NOTICE THEY ARE NAMED DIFFERENT


##### HOW MANY MCI, DEMOGRPHICS, YADAYDA

### EXPORT -- SET WITH RELATIVE PATH


```{r exporting}



```


